---
title: "Lab4"
author: "Oskar Hidén - oskhi827"
date: "10/18/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Part 1 - Implementing GP Regression
```{r}
#install.packages('kernlab')
library(kernlab)

# Computing the whole covariance matrix K from the kernel.
cov_function = function(X, Xstar){
  return(kernelMatrix(kernel = SEkernel, x=X, y=Xstar)) 
}

# Algorithm 2.1 sunns in n^3/6 instead of O(n^3). GP can be estimated faster, O(n) with eg. KISS-GP
posteriorGP = function(X_input, y_targets, k_cov_function, sigmaNoise=1, XStar){
  A = k_cov_function(X_input, X_input)
  A = A + diag(length(X_input))*sigmaNoise^2
  L = t(chol(A)) # chol Returns t(L)
  L_y = solve(L, y_targets)
  alpha = solve(t(L), L_y)
  
  k_star = k_cov_function(X_input, XStar)
  f_star = t(k_star)%*%alpha
  
  v = solve(L,k_star)
  V_f_star = k_cov_function(XStar, XStar) - t(v)%*%v
  return(list("mean"=f_star, "cov" = V_f_star))
}

sigma_f = 1
ell = 0.3
SEkernel <- rbfdot(sigma = 1/(2*ell^2)) # Reparametrize the rbfdot (which is the SE kernel) in kernlab.
sigma_n = 0.1

# 2 - Update prior with a signle observation
x=0.4
y=0.719
xGrid <- seq(-1,1,length=20) # x_star

posterior = posteriorGP(x, y, cov_function, sigma_n, xGrid)

plot(xGrid, posterior$mean, ylim = c(-2.5,2.5)) # posterior mean
std_dev = sqrt(diag(posterior$cov))
points(xGrid, posterior$mean + 1.96*std_dev, col="blue")
points(xGrid, posterior$mean - 1.96*std_dev, col="blue")

# 3 - Update posterior with another point. (by recalculating posterior wiht Algorithm 2.1)
x = c(0.4, -0.6)
y = c(0.719 , -0.044)
posterior = posteriorGP(x, y, cov_function, sigma_n, xGrid)

plot(xGrid, posterior$mean, ylim = c(-2.5,2.5)) # posterior mean
std_dev = sqrt(diag(posterior$cov))
points(xGrid, posterior$mean + 1.96*std_dev, col="blue")
points(xGrid, posterior$mean - 1.96*std_dev, col="blue")


# 4 - Calculate the posterior with five observations.
x = c(-1.0, -0.6, -0.2, 0.4, 0.8)
y = c(0.768, -0.044, -0.940, 0.719, -0.664)
posterior = posteriorGP(x, y, cov_function, sigma_n, xGrid)

plot(xGrid, posterior$mean, ylim = c(-2.5,2.5)) # posterior mean
std_dev = sqrt((diag(posterior$cov)))
points(xGrid, posterior$mean + 1.96*std_dev, col="blue")
points(xGrid, posterior$mean - 1.96*std_dev, col="blue")

# 5 - Redo (4) with the same sigma_f but with a higher ell.
sigma_f = 1
ell = 1
SEkernel <- rbfdot(sigma = 1/(2*ell^2))

x = c(-1.0, -0.6, -0.2, 0.4, 0.8)
y = c(0.768, -0.044, -0.940, 0.719, -0.664)
posterior = posteriorGP(x, y, cov_function, sigma_n, xGrid)

plot(xGrid, posterior$mean, ylim = c(-2.5,2.5)) # posterior mean
std_dev = sqrt((diag(posterior$cov)))
points(xGrid, posterior$mean + 1.96*std_dev, col="blue")
points(xGrid, posterior$mean - 1.96*std_dev, col="blue")

```

The two plots has the same sigma_f but different l-values. The last plot is smoother, with a higher l value. And the 95% confidence intervall is closer to posterior mean. 


## Part 2 - GP Regression with kernlab
```{r}
temprature = read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/TempTullinge.csv",header=TRUE,sep=";")
time = (1:2190)
nr_year = 2190/365
day = rep((1:365), nr_year)

reduce_data = function(array , nth){
  array[seq(1, length(array), nth)]
}

temp = reduce_data(temprature$temp,5)
time = reduce_data(time, 5)
day = reduce_data(day, 5)

set_se_kernel = function(ell, sigma_f){
  se_kernel = function(x , y){
    r_square = sum((x-y)*(x-y)) # Euclidian distance^2
    return(sigma_f^2*exp(-r_square/(2*ell^2)))
  }
  class(se_kernel) <- "kernel"
  return(se_kernel)
}

se_kernel = set_se_kernel(ell=1, sigma_f=1)

# Evaluate kernel in x=1 and x'=2
se_kernel(1,2)

k = function(x, x_star){
  return(kernelMatrix(kernel = se_kernel, x=x, y=x_star))
}

# Evaluate the covariance matrix, k
x = c(1, 3, 4)
x_star = c(2, 3, 4)
k(x, x_star)


# 2 - Find f* with function: gausspr(kernlab)
sigma_f = 20
ell = 0.2
se_kernel = set_se_kernel(ell=ell, sigma_f=sigma_f)

# Find sigma_n
quadratic_reg = lm(temp ~ time + I(time^2))
sigma_n = sd(quadratic_reg$residuals)

gp_model = gausspr(x = time, y = temp, type="regression", kernel=se_kernel, var = sigma_n^2, variance.model=TRUE)

mean_pred = predict(gp_model, time)
plot(time, temp)
points(time, mean_pred, type = "l", col="red")
#lines(time, mean_pred + 1.96 * predict(gp_model, time, type="sdeviation")) # ---- wrong due to scaleing done by gausspr....(bugg)


# 3 - Scale and calculate with Algorithm 2.1
mean_temp = mean(temp)
sd_temp = sd(temp)
time_scale = scale(time)
temp_scale = scale(temp)

# Find sigma_n
quadratic_reg = lm(temp_scale ~ time_scale + I(time_scale^2))
sigma_n = sd(quadratic_reg$residuals)

post_gp = posteriorGP(time_scale, temp_scale, k, sigma_n, time_scale)
st_dev_new = sqrt(diag(post_gp$cov))*sd_temp
mean_pred_new = post_gp$mean*sd_temp+mean_temp
plot(time, temp)
lines(time, mean_pred_new, col ="red")
lines(time, mean_pred_new+1.96*st_dev_new, col="blue")
lines(time, mean_pred_new-1.96*st_dev_new, col="blue")


# 4 - use of Day
day_scale = scale(day)

#Find sigma_n
quadratic_reg = lm(temp_scale ~ day_scale + I(day_scale^2))
sigma_n = sd(quadratic_reg$residuals)

post_gp = posteriorGP(day_scale, temp_scale, k, sigma_n, day_scale)
st_dev_day = sqrt(diag(post_gp$cov))*sd_temp
mean_pred_day = post_gp$mean*sd_temp+mean_temp

# old plot
plot(time, temp)
lines(time, mean_pred_new, col ="red")
lines(time, mean_pred_new+1.96*st_dev_new, col="blue")
lines(time, mean_pred_new-1.96*st_dev_new, col="blue")
# new lines from day
lines(time, mean_pred_day, col ="orange")
lines(time, mean_pred_day+1.96*st_dev_day, col="green")
lines(time, mean_pred_day-1.96*st_dev_day, col="green")
```
Pros and cons of each models?

```{r}
# 5 - Generalization periodic kernel
sigma_f = 20
ell_1 = 1
ell_2 = 10
d = 365/sd(time)

set_gen_kernel = function(ell_1, ell_2, sigma_f, d){
  gen_kernel = function(x , y){
    r_square = sum((x-y)*(x-y)) #euclidian distance ^2
    return(sigma_f^2* exp(-sin(pi*sqrt(r_square)/d)/(ell_1^2)) * exp(-r_square/(2*ell_2^2)))
  }
  class(gen_kernel) <- "kernel"
  return(gen_kernel)
}

k_gen = function(x, x_star){
  return(kernelMatrix(kernel = gen_kernel, x=x, y=x_star))
}

# Set sigma_n
quadratic_reg = lm(temp ~ time + I(time^2))
sigma_n = sd(quadratic_reg$residuals)

gen_kernel = set_gen_kernel(ell_1, ell_2, sigma_f, d)
gp_model = gausspr(x = time, y = temp, type="regression", kernel=gen_kernel, var = sigma_n^2, variance.model=TRUE)

mean_pred_gen = predict(gp_model, time)
plot(time, temp)
points(time, mean_pred, type="l", col="red")
points(time, mean_pred_day, type = "l", col="blue")
points(time, mean_pred_gen, type = "l", col="green")

```
Discuss the results:
When predicting using time and sqare exponential kernel we have a low correlation between the same date between two consecutive years. But when using day and square exponential kernel we assume the same correlation between the same date, no mather which year that dat is from. And we get a prediction that does not take the long term change (ex. warmer winters) into account. Day is not that smooth when we change year, because day=1 and day=365 will not get a strong correlation, even tho they are close.

With the periodic kernel we give two ecual dates a stronger correlation if they are in two consecutive years. We also mannage to capture the long term trend better, because of the second exp() of the kernel.


## Part 3 - GP Classification with kernlab
```{r}
#install.packages("AtmRay") # To make 2D grid like in Matlab's meshgrid.
library(AtmRay)

data <- read.csv("https://github.com/STIMALiU/AdvMLCourse/raw/master/GaussianProcess/Code/banknoteFraud.csv", header=FALSE, sep=",")
names(data) <- c("varWave","skewWave","kurtWave","entropyWave","fraud")
data[,5] <- as.factor(data[,5])

set.seed(111)
SelectTraining <- sample(1:dim(data)[1], size = 1000,replace = FALSE)
train = data[SelectTraining,]
gp_fraud = gausspr(fraud ~ varWave + skewWave, data=train, type="classification")

fraud_pred = predict(gp_fraud, train[,1:4])
# Confusion matrix
table(fraud_pred, train[,5])
```
